### 2020-11-25 latex/acrobot.tex
New proof of oscillation energy gain, needs edits next

\begin{thm}\label{thm:acrobot-energy-stabilization}
    Consider the simple acrobot ~\eqref{eqn:acrobot-hamiltonian} constrained by
    the VNHC ~\eqref{eqn:acrobot-constraint}, whose
    constraint manifold is \(\Gamma \simeq \SxR\).
    Let 
    \[
        E(q_u,p_u) := \frac{p_u^2}{10ml^2} + 3mgl(1 - \cos(q_u))
        ,
    \]
    be the energy of the simple pendulum obtained by setting \(I = 0\).
\begin{enumerate}
    \item For all \(\epsilon > 0\), there exists \(I > 0\) small enough that 
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on the set
    \[
        \mathcal{O}_\epsilon := \left\{(q_u,p_u) \in \Gamma 
        \mid E(\epsilon,0) < E(q_u,p_u) < E(\pi,0) \right\}
        .
    \]
    If instead \(I < 0\), the VNHC dissipates energy on \(\mathcal{O}_\epsilon\).
\item Define \(b : \SxR_{> 0} \rightarrow \R\) by
    \[
        b(\beta,\rho_0) := 
        \frac{5m^2 g l^3 \bar{q}_a \left(
            m^2gl^3\left(18s_\beta^2 + 30c_\beta(1 - c_\beta)\right)
            - c_\beta\rho_0^2
        \right)}{
        |\rho_0|\sqrt{\rho_0^2 - 30m^2gl^3(1 - c_\beta)}
        }
        ,
    \]
    and fix \(\bar{\rho} > \sqrt{60m^2gl^3}\).
    Suppose \(\int \limits_{0}^{2\pi} b(\sigma,\rho_0)d\sigma > 0\) for all 
    \(\rho_0 \in \, \left]\sqrt{60m^2gl^3}, \bar{\rho}\right]\).
    Then for all \(\epsilon > 0\) there exists \(I > 0\) small enough that
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on
    \[
        \Omega_\epsilon := \left\{(q_u,p_u) \in \Gamma 
        \mid E(\epsilon,0) < E(q_u,p_u) < E(0,\bar{\rho})\right\}
        .
    \]
    If instead \(I < 0\), the VNHC dissipates energy on \(\Omega_\epsilon\).
\end{enumerate}
\end{thm}

@@

Now consider the solution \(x(t,\eta_0,I)\) to ~\eqref{eqn:khalil-setup}.
Performing a first-order Taylor series expansion at \(I = 0\), we get
\(x(t,\eta_0,I) = x_0(t,\eta_0) + I x_1(t,\eta_0) + R(t,\eta_0,I)\), where the
remainder term \(R(t,\eta_0,I)\) is smooth and \(O(I^2)\), \ie,
\[
    \lim \limits_{I \to 0} \frac{R(t,I)}{I} = 0
    .
\]
One can show \(x_1(t,\eta_0)\) is the solution to the scalar time-varying ODE
\begin{equation}\label{eqn:khalil-perturbation-firstorder}
    \begin{cases}
        \dot{x}_1 = \pdiff{f}{x}(t,x_0(t,\eta_0),0)x_1 + \pdiff{f}{I}(t,x_0(t,\eta_0),0)
        , \\
        x_1(t_0,\eta_0) = 0
        .
    \end{cases}
\end{equation}

@@

Theorem \ref{thm:khalil-perturbation} tells us that we can approximate the
solution of the nonlinear system by the Taylor approximation 
\(x(t,\eta_0,I) \approx x_0(t,\eta_0) + I x_1(t,\eta_0)\).
When \(I\) is small enough, solutions of the nonlinear system and
the Taylor approximation are the same up to order \(I^2\) along compact time
intervals.
Since we know that the acrobot behaves like a pendulum at \(I = 0\), we can
use this theory to prove energy injection properties on the acrobot by studying
the simple pendulum.

@@

@@

The Poincar\'{e} section \(P_\mathcal{O}\) allows us understand the evolution of
\(\hat{\mu}(\alpha,\mu_0,I)\) by studying the evolution of a discrete time system;
abusing notation, we set the state of this discrete time system to be
\(\hat{\mu}(n)\) for \(n \in \mathbb{Z}_{\geq 0}\). 
The dynamics of \(\hat{\mu}(n)\) are
\begin{equation}\label{eqn:muhat-discrete}
    \begin{cases}
        \hat{\mu}(n+1) = P_\mathcal{O}(\hat{\mu}(n)) 
        = \hat{\mu}(n) + I K Q(\hat{\mu}(n)) + R(\pi,\hat{\mu}(n),I)
        , \\
        \hat{\mu}(0) = \mu_0
        .
    \end{cases}
\end{equation}

To show energy gain on \(\mathcal{O}\), it is sufficient to 
show this discrete time system is non-decreasing.
That is, we wish to show there exists \(I^\star \in ]0, I_1]\) where, for all 
\(\delta > 0\) and \(\mu_0 \in ]0,\pi-\delta]\), there exists \(N > 0\) so that
for all \(n \geq N\), \(\hat{\mu}(n) \notin ]0,\pi-\delta]\).

% TODO: Replace from here
This fact, coupled with knowledge that \(R(\pi,\mu_0,I)\) is \(O(I^2)\), 
implies there exists \(I > 0\) small enough that
\[
    P_\mathcal{O}(\mu_0) = \mu_0 + IKQ(\mu_0) + R(\pi,\mu_0,I) > 0
    .
\]
Unfortunately, this \(I\) may not be the
same for all \(\mu_0 \in \, ]0,\pi[\), since
the value of \(R(\pi,\mu_0,I)\) could become more negative
for other initial conditions.
We deal with this issue by taking the compact set
\[
    \bar{\mathcal{O}} := \left\{(q_u,p_u) \in \Gamma 
    \mid E(q_u,p_u) \leq E(\pi,0) \right\}
    .
\]
The remainder term \(R(\alpha,\mu_0,I)\) is smooth and of order \(I^2\),
which means it can be written in the form 
\(R(\alpha,\mu_0,I) = I^2\tilde{R}(\alpha,\mu_0,I)\)
for some \(\tilde{R}\) a smooth function.
Since \(\bar{\mathcal{O}}\) is compact, \(\tilde{R}\) attains its minimum
\({r}\) somewhere on \(\bar{\mathcal{O}}\).
Hence, \(R(\alpha,\mu_0,I)\) is bounded below on \(\mathcal{O}\) by
\(I^2{r}\), which implies 
\begin{equation}\label{eqn:acrobot-Po-lowerbound}
    P_\mathcal{O}(\mu_0) \geq \mu_0 + IKQ(\mu_0) + I^2{r}
    .
\end{equation}
This equation further implies that, for fixed \(I\),
\(P_\mathcal{O}(\mu_0)\) is monotonically increasing in \(\mu_0\).
Thus, for every \(\epsilon > 0\) and \(\mu_0 \in \, [\epsilon,\pi[\) we have
\[
    P_\mathcal{O}(\mu_0) \geq P_\mathcal{O}(\epsilon)
    = \epsilon + IKQ(\epsilon) + I^2{r}
    .
\]
Since \(I\) is assumed to be small, \(I^2\) shrinks rapidly.
Therefore, we can always find a small enough \(I > 0\) making 
\(\epsilon + I^2{r} \geq 0\), so that
\[
    P_\mathcal{O}(\epsilon) \geq IKQ(\epsilon) > 0
    .
\]

We conclude that, for all \(m\), \(g\), \(l\), \(\bar{q}_a\), and for
any \(\epsilon > 0\), there exists \(I^\prime(\epsilon)\) small enough so that,
for all \(0 < I \leq I^\prime(\epsilon)\),
the Poincar\'{e} section \(P_\mathcal{O}(\mu_0)\) is positive for all
\(\mu_0 \in [\epsilon,\pi[\) and 
\(\lim \limits_{\mu_0 \to \pi} P_\mathcal{O}(\mu_0) = \infty\).

\subsection*{Energy gain on \(\mathcal{O}_\epsilon\)}
To simplify notation, denote by \(x(t) := (q_u(t),p_u(t))\) the orbit of the
acrobot with initial condition \(x(0)\).
Let \(\epsilon > 0\) and define
\[
    \mathcal{O}_\epsilon := \left\{(q_u,p_u) \in \Gamma
    \mid E(\epsilon,0) < E(q_u,p_u) < E(\pi,0)\right\}
    ,
\]
with \(\bar{\mathcal{O}}_\epsilon\) its (compact) closure.

Let \(x(0) \in \mathcal{O}_\epsilon \subset \mathcal{O}\).
Since \(\dot{\alpha} > 0\), the orbit \(x(t)\) spirals clockwise around
the origin.
If this orbit remains in \(\mathcal{O}\), it must intersect
the \(q_u\) axis at some radius \(\mu = \delta\)
which depends on the initial condition \(x(0)\).
Take 
\[
    \underbar{\delta} = 
    \min\limits_{x(0) \in \bar{\mathcal{O}}_\epsilon} \delta(x(0))
    ,
\] 
which exists because \(\bar{\mathcal{O}}_\epsilon\) is compact.
Furthermore, \(\underbar{\delta} \leq \epsilon\) because
\(x(0) = (\epsilon,0) \in \bar{\mathcal{O}}_\epsilon\).
From the previous perturbation analysis,
there exists \(I^\prime(\underbar{\delta})\) small enough that
\(P_\mathcal{O}(\mu_0)\) is positive for all 
\(\mu_0 \in [\underbar{\delta},\pi[\).

Suppose \(x(0) \in \mathcal{O}_\epsilon\) and that the orbit
intersects the \(q_u\)-axis at some point \(x(t_0) = (u_0, 0)\), which
has pseudo-radius \(\mu = \abs{u_0}\).
The Poincar\'{e} section \(P_\mathcal{O}(\abs{u_0})\) is positive because 
\(\abs{u_0} \geq \underbar{\delta}\).
As the acrobot continues its oscillation, it will intersect the \(q_u\) axis
again at a new point \(x(t_1) = (u_1,0)\).

Setting \(p_1 := P_\mathcal{O}(\abs{u_0})\),
Theorem \ref{thm:khalil-perturbation} guarantees that,
for all \(I \in ]0,I^\prime(\underbar{\delta})]\),
\[
    \abs{\abs{u_1} - p_1} \leq cI^2
    ,
\]
where \(c > 0\) is an independent constant.
This implies that \(\abs{u_1} \geq p_1 - cI^2\). 
We also know that
\[
    p_1 - \abs{u_0} \geq IKQ(\abs{u_0}) \geq IKQ(\underbar{\delta}) > 0
    .
\]
There exists \(I^\star \in ]0,I^\prime(\underbar{\delta})]\) where
\(c{I^\star}^2 < I^\star KQ(\underbar{\delta})\), which means
\begin{align*}
    \abs{u_1} - \abs{u_0} 
    &= \left(\abs{u_1} - p_1\right) + \left(p_1 - \abs{u_0}\right)
    \\
    &\geq \left(p_1 - c{I^\star}^2 - p_1\right) + I^\star KQ(\underbar{\delta})
    \\
    &= I^\star KQ(\underbar{\delta}) - c{I^\star}^2
    \\
    &> 0
    .
\end{align*}

Since \(u_0 \in [\underbar{\delta},\pi[\) was arbitrary, 
we can apply Theorem \ref{thm:khalil-perturbation} iteratively:
initializing the acrobot at \(x(t_i) = (u_i, 0)\), the
next intersection with the \(q_u\) axis will be at
\(x(t_{i+1}) = (u_{i+1},0)\) with \(\abs{u_{i+1}} > \abs{u_i}\).

Because the Poincar\'{e} section is monotonically increasing, 
the orbit will eventually reach an intersection point \(x(t_N) = (u_N,0)\)
such that \(P_\mathcal{O}(\abs{u_N}) > \pi + c{I^\star}^2\). 
At this point, the orbit will no longer return to the
\(q_u\)-axis.
Instead, it will escape \(\mathcal{O}\) entirely, which means
it also escapes \(\mathcal{O}_\epsilon\).
Hence, the orbit must escape any compact subset of 
\(\mathcal{O}_\epsilon\) in finite time. 
We conclude that the acrobot is gaining energy on \(\mathcal{O}_\epsilon\).

Setting \(I < 0\), the Poincar\'{e} section \(P_\mathcal{O}(\mu_0)\) is now
monotonically decreasing in \(\mu_0\).
By the same arguments presented in this section, the constraint 
must be dissipating energy on \(\mathcal{O}_\epsilon\).

@@

\subsection*{Perturbation Analysis for Rotations}
The rotation region for the nominal pendulum is the set
\[
    \mathcal{R} := \left\{ (q_u,p_u) \in \Gamma \mid E(q_u,p_u) > E(\pi,0)\right\}
    .
\]
Figure \ref{fig:pendulum-level-sets} shows that orbits of the 
pendulum on \(\mathcal{R}\) look like open curves.
Performing a similar process to what we did for oscillations,
we wish to find a new set of coordinates \((\beta,\rho)\)
where \(\rho\) remains constant on level sets of \(E\)
and \(\beta \in \Sone\) is always increasing on \(\mathcal{R}\).

>>>>

\begin{thm}\label{thm:acrobot-energy-stabilization}
    Consider the simple acrobot ~\eqref{eqn:acrobot-hamiltonian} constrained by
    the VNHC ~\eqref{eqn:acrobot-constraint}, whose
    constraint manifold is \(\Gamma \simeq \SxR\).
    Let 
    \[
        E(q_u,p_u) := \frac{p_u^2}{10ml^2} + 3mgl(1 - \cos(q_u))
        ,
    \]
    be the energy of the simple pendulum obtained by setting \(I = 0\).
\begin{enumerate}
    \item For all \(\epsilon > 0\), there exists \(I > 0\) small enough that 
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on
    \[
        \mathcal{O} := \left\{(q_u,p_u) \in \Gamma 
        \mid E(q_u,p_u) < E(\pi,0) \right\}
        .
    \]
    If instead \(I < 0\), the VNHC dissipates energy on \(\mathcal{O}\).
\item Define \(b : \SxR_{> 0} \rightarrow \R\) by
    \[
        b(\beta,\rho_0) := 
        \frac{5m^2 g l^3 \bar{q}_a \left(
            m^2gl^3\left(18s_\beta^2 + 30c_\beta(1 - c_\beta)\right)
            - c_\beta\rho_0^2
        \right)}{
        |\rho_0|\sqrt{\rho_0^2 - 30m^2gl^3(1 - c_\beta)}
        }
        ,
    \]
    and fix \(\bar{\rho} > \sqrt{60m^2gl^3}\).
    Suppose \(\int \limits_{0}^{2\pi} b(\sigma,\rho_0)d\sigma > 0\) for all 
    \(\rho_0 \in \, \left]\sqrt{60m^2gl^3}, \bar{\rho}\right]\).
    Then there exists \(I > 0\) small enough that
    ~\eqref{eqn:acrobot-constraint} injects energy into the acrobot on
    \[
        \Omega := \left\{(q_u,p_u) \in \Gamma 
        \mid E(q_u,p_u) < E(0,\bar{\rho})\right\}
        .
    \]
    If instead \(I < 0\), the VNHC dissipates energy on \(\Omega\).
\end{enumerate}
\end{thm}

@@

Now consider the solution \(x(t,\eta_0,I)\) to ~\eqref{eqn:khalil-setup}.
Performing a first-order Taylor series expansion at \(I = 0\), we get
\(x(t,\eta_0,I) = x_0(t,\eta_0) + I x_1(t,\eta_0) + R(t,\eta_0,I)\), where the
remainder term \(R(t,\eta_0,I)\) is smooth and \(O(I^2)\), \ie,
there exists \(c > 0\) such that 
\[
    \norm{ R(t,\eta_0,I) } \leq c \abs{I^2}
    ,
\]
for all \((t,\eta_0,I) \in [t_0,t_1] \times D \times [-I_0,I_0]\), and
\[
    \lim \limits_{I \to 0} \frac{R(t,\eta_0,I)}{I} = 0
    .
\]
One can show \(x_1(t,\eta_0)\) is the solution to the scalar time-varying ODE
\begin{equation}\label{eqn:khalil-perturbation-firstorder}
    \begin{cases}
        \dot{x}_1 = \pdiff{f}{x}(t,x_0(t,\eta_0),0)x_1 + \pdiff{f}{I}(t,x_0(t,\eta_0),0)
        , \\
        x_1(t_0,\eta_0) = 0
        .
    \end{cases}
\end{equation}

@@

Theorem \ref{thm:khalil-perturbation} tells us that we can approximate the
solution of the nonlinear system by the Taylor approximation 
\(x(t,\eta_0,I) \approx x_0(t,\eta_0) + I x_1(t,\eta_0)\).
When \(I\) is small enough, solutions of the nonlinear system and
the Taylor approximation are the same up to order \(O(I^2)\) along compact time
intervals.
Since we know that the acrobot behaves like a pendulum at \(I = 0\), we can
use this theory to prove energy injection properties on the acrobot by studying
the simple pendulum.

@@

This fact, coupled with knowledge that \(R(\pi,\mu_0,I)\) is \(O(I^2)\), 
implies there exists \(I > 0\) small enough that
\[
    P_\mathcal{O}(\mu_0) = \mu_0 + IKQ(\mu_0) + R(\pi,\mu_0,I) > 0
    .
\]
Unfortunately, perturbation analysis does not guarantee 
that \(I\) is the same for all \(\mu_0 \in \, ]0,\pi[\), since the value of
\(R(\pi,\mu_0,I)\) could become more negative for other initial conditions.
We will deal with this next.

@@

The Poincar\'{e} section \(P_\mathcal{O}\) allows us understand the evolution of
\(\hat{\mu}(\alpha,\mu_0,I)\) by studying the evolution of a discrete time system;
abusing notation, we set the state of this discrete time system to be
\(\mu(n)\) for \(n \in \mathbb{Z}_{\geq 0}\). 
The dynamics of \(\mu(n)\) are
\begin{equation}\label{eqn:muhat-discrete}
    \begin{cases}
        \mu(n+1) = P_\mathcal{O}(\mu(n)) 
        = \mu(n) + I K Q(\mu(n)) + R(\pi,\mu(n),I)
        , \\
        \mu(0) = \mu_0
        .
    \end{cases}
\end{equation}

To show energy gain on \(\mathcal{O}\), it is sufficient to 
show this discrete time system is strictly increasing.
This is because \(\mu(n)\) represents the distance along the \(q_u\)-axis
when an orbit intersects the \(q_u\)-axis for the \(n\)th time, assuming
the orbit was initialized at \((q_u,p_u) = (\mu_0,0)\).

Formally, we wish to show there exists \(I^\star \in ]0, I_1]\) where, for all 
\(\delta > 0\) and \(\mu_0 \in ]0,\pi-\delta]\), there exists \(N > 0\) so that
for all \(n \geq N\), \(\mu(n) \notin ]0,\pi-\delta]\).

Linearizing ~\eqref{eqn:muhat-discrete} at \(\mu = 0\) yields
\[
    P^\prime_\mathcal{O}(0) = 1 + I K Q^\prime(0) +
    \pdiff{R}{\mu}(\pi,0,I)
    .
\]
If there is some value of \(I \in ]0,I_1]\) for which this term is greater than
\(1\), then \(0\) is a repeller of the discrete time system.
To show such a value of \(I\) exists, we first compute 
\[
    Q^\prime(0) = \int \limits_0^\pi \pdiff{b}{\mu}(\sigma,0) d\sigma
    .
\]
MATLAB reveals that 
\[
    \lim \limits_{\mu_0 \to 0^+}
    \pdiff{b}{\mu}(\alpha,\mu_0) 
    = -\frac{\sqrt{2}}{2} \left(11\sin(\alpha)^2 - 6\right)
    ,
\]
which means 
\[
    Q^\prime(0)  = \frac{\pi}{2\sqrt{2}} > 0
    .
\]
Since the remainder term \(R(\alpha,\mu_0,I)\) is \(O(I^2)\), 
its partial derivative \(\pdiff{R}{\mu}(\pi,\mu_0,I)\) is also \(O(I^2)\).
Hence, it can be written in the form
\(I^2 \tilde{R}^\prime(\mu_0,I)\) where \(\tilde{R}^\prime(\mu_0,I)\) is smooth and
zero at \(I = 0\).
Since \(I\) is assumed to be small, \(I^2\) shrinks rapidly;
thus, there exists \(I_2 \in ]0,I_1]\) where 
\(I_2 K Q^\prime(0) + I_2^2 \tilde{R}^\prime(0,I) > 0\).
Hence,
\[
    P^\prime_\mathcal{O}(0) \geq 1 + IKQ^\prime(0) + I^2 \tilde{R}^\prime(0,I) 
    > 1
    ,
\] 
for all \(I \in ]0, I_2]\), .

This proves \(0\) is a repeller of the discrete time system, which means
there exists some (unknown) \(\epsilon > 0\) where the interval
\(]0,\epsilon[\) is negatively invariant for ~\eqref{eqn:muhat-discrete}.
What's more, all solutions starting in this interval will flow towards the value
\(\mu = \epsilon\).
In other words, we have found a value \(I_2 > 0\) small enough
that \(\mu(n)\) is strictly increasing on \([0,\epsilon[\).

Next, recall that \(R(\pi,\mu_0,I)\) is smooth and \(O(I^2)\)
and is therefore bounded below on \(\mu \in [\epsilon,\pi]\) by some value
\(-I^2r\).
Furhtermore, \(Q(\mu_0)\) is a strictly increasing function, so
for\(\mu_0 \in [\epsilon,\pi[\) we have
\[
    P_\mathcal{O}(\mu_0) \geq \mu_0 + I K Q(\epsilon) - I^2 r
    .
\]
There clearly exists a small enough \(I^\star \in ]0,I_2]\) making
\(I^\star K Q(\epsilon) - (I^\star)^2 r > 0\), which means
\[
    P_\mathcal{O}(\mu_0) > \mu_0
    ,
\]
for all \(\mu_0 \in [\epsilon,\pi[\).

We have thus shown that \(P_\mathcal{O}(\mu_0) > \mu_0\) for all
\(\mu_0 \in ]0,\pi[\).
Looking at ~\eqref{eqn:muhat-discrete}, this means
\(\mu(n+1) > \mu(n)\) and all solutions of the discrete time system will flow
towards \(\mu = \pi\).
Hence, orbits of the constrained dynamics will eventually escape
\(\mathcal{O}\).

We conclude that, for all \(m\), \(g\), \(l\), \(\bar{q}_a\), 
there exists \(I > 0)\) small enough that the constraint 
~\eqref{eqn:acrobot-constraint} is injecting energy on \(\mathcal{O}\).

Setting \(I < 0\), the Poincar\'{e} section \(P_\mathcal{O}(\mu_0)\) is now
monotonically decreasing in \(\mu_0\).
By the same arguments presented in this section, the constraint 
must be dissipating energy on \(\mathcal{O}\).

@@

\subsection{Perturbation Analysis for Rotations}
The rotation region for the nominal pendulum is the set
\[
    \mathcal{R} := \left\{ (q_u,p_u) \in \Gamma \mid E(q_u,p_u) > E(\pi,0)\right\}
    .
\]
Figure \ref{fig:pendulum-level-sets} shows that orbits of the 
pendulum on \(\mathcal{R}\) look like open curves.
Performing a similar process to what we did for oscillations,
we wish to find a new set of coordinates \((\beta,\rho)\)
where \(\rho\) remains constant on level sets of \(E\)
and \(\beta \in \Sone\) is always increasing on \(\mathcal{R}\).
